### 2017-03-19 paper/arrows.tex
wip on paper. added pictures from the presentation.

\section{Arrows}
Arrows were introduced by Hughes \citHughes ~as a general interface for computation. An arrow \code{arr a b} can be look at as a computation that converts an input \code{a} to an output \code{b}. This is defined in the arrow typeclass:
\begin{lstlisting}[frame=htrbl]
class Arrow arr where
	arr :: (a -> b) -> arr a b
	(>>>) :: arr a b -> arr b c -> arr a c
	first :: arr a b -> arr (a,c) (b,c)
\end{lstlisting}
\lstinline{arr} is used to lift an ordinary function to an arrow type. This can be thought of as analogous to the monadic \lstinline{return}. The \lstinline{>>>} operator, in a similar way, is analogous to the monadic composition operator \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}. And lastly, the \lstinline{first} operator, which takes the input arrow from b to c and converts it into an arrow on pairs with the second argument untouched, is also needed for actual useful code as without it, we wouldn't have a way to save input across arrows.
\\\\
The most prominent instances of this interface are regular functions \lstinline{(->)},
\begin{lstlisting}[frame=htrbl]
instance Arrow (->) where
	arr f = f
	f >>> g = g . f
	first f = \(a, c) -> (f a, c) 
\end{lstlisting}
and the Kleisli type.
\begin{lstlisting}[frame=htrbl]
instance Arrow (Kleisli m) where
	arr f = Kleisli $ return . f
	f >>> g = Kleisli $ \a -> f a >>= g
	first f = Kleisli $ \(a,c) -> f a >>= \b -> return (b,c)
\end{lstlisting}
With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \lstinline{first}, called \lstinline{second},
\begin{lstlisting}[frame=htrbl]
second :: Arrow arr => arr a b -> arr (c, a) (c, b)
second f = arr swap >>> first f >>> arr swap
	where swap (x, y) = (y, x)
\end{lstlisting}
the *** combinator which combines first and second to handle two inputs in one arrow,
\begin{lstlisting}[frame=htrbl]
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{lstlisting}
and the \&\&\& combinator that constructs an arrow which outputs 2 different values like ***, but takes only one input.
\begin{lstlisting}[frame=htrbl]
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{lstlisting}
A short example given by Hughes on how to use this is \lstinline{add} over arrows:
\begin{lstlisting}[frame=htrbl]
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{lstlisting}
The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
\\\\
\textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used. We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}

>>>>

\section{Arrows}
\begin{center}
	\includegraphics[]{images/arrow.png}
\end{center}
Arrows were introduced by Hughes \citHughes ~as a general interface for computation. An arrow \code{arr a b} can be look at as a computation that converts an input \code{a} to an output \code{b}. This is defined in the arrow typeclass:

\begin{minipage}{\textwidth}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting}[frame=htrbl]
class Arrow arr where
arr :: (a -> b) -> arr a b

(>>>) :: arr a b -> arr b c -> arr a c

first :: arr a b -> arr (a,c) (b,c)
\end{lstlisting}
\end{minipage}
~~~~
\begin{minipage}{0.25\textwidth}
	~\\~\\~\\
	\includegraphics[scale=0.6]{images/arr}~\\~\\
	\includegraphics[scale=0.6]{images/compose}~\\~\\
	\includegraphics[scale=0.6]{images/first}~\\~\\
\end{minipage}
\end{minipage}
\lstinline{arr} is used to lift an ordinary function to an arrow type. This can be thought of as analogous to the monadic \lstinline{return}. The \lstinline{>>>} operator, in a similar way, is analogous to the monadic composition operator \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}. And lastly, the \lstinline{first} operator, which takes the input arrow from b to c and converts it into an arrow on pairs with the second argument untouched, is also needed for actual useful code as without it, we wouldn't have a way to save input across arrows.
\\\\
The most prominent instances of this interface are regular functions \lstinline{(->)},
\begin{lstlisting}[frame=htrbl]
instance Arrow (->) where
	arr f = f
	f >>> g = g . f
	first f = \(a, c) -> (f a, c) 
\end{lstlisting}
and the Kleisli type
\begin{lstlisting}[frame=htrbl]
data Kleisli m a b = Kleisli { run :: a -> m b }
\end{lstlisting}
as well:
\begin{lstlisting}[frame=htrbl]
instance Monad m => Arrow (Kleisli m) where
	arr f = Kleisli $ return . f
	f >>> g = Kleisli $ \a -> f a >>= g
	first f = Kleisli $ \(a,c) -> f a >>= \b -> return (b,c)
\end{lstlisting}
\begin{figure}
	\centering
	\begin{tabular}{cc}
		\subcaptionbox{first\label{1}}{\includegraphics[width = 1.5in]{images/first}} &
		\subcaptionbox{second\label{2}}{\includegraphics[width = 1.5in]{images/second}} \\
		\subcaptionbox{(***)\label{1}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		\subcaptionbox{(\&\&\&)\label{2}}{\includegraphics[width = 1.5in]{images/andandand}}\
	\end{tabular}
	\caption{syntactic sugar}
	\label{3}
\end{figure}
With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \lstinline{first}, called \lstinline{second},
\begin{lstlisting}[frame=htrbl]
second :: Arrow arr => arr a b -> arr (c, a) (c, b)
second f = arr swap >>> first f >>> arr swap
	where swap (x, y) = (y, x)
\end{lstlisting}
the *** combinator which combines first and second to handle two inputs in one arrow,
\begin{lstlisting}[frame=htrbl]
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{lstlisting}
and the \&\&\& combinator that constructs an arrow which outputs 2 different values like ***, but takes only one input.
\begin{lstlisting}[frame=htrbl]
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{lstlisting}
A short example given by Hughes on how to use this is \lstinline{add} over arrows:
\begin{lstlisting}[frame=htrbl]
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{lstlisting}
The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
\\\\
\textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used. We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}

### 2017-03-19 paper/basicSkeletons.tex
wip on paper. added pictures from the presentation.

\section{Map-based Skeletons}
With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.

\fixme{Extra Kapitel für die Basic Skeletons, vll den Syntactic Sugar dann danach machen statt erst nach den maps}

@@

\subsection{parMap}
\begin{center}
	\includegraphics[scale=0.7]{images/parMap}
\end{center}
\code{parMap} is probably the most common skeleton for parallel programs. We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}. 
Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf, ArrowApply arr) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f =
	(arr $ \as -> (f, as)) >>>
	(first $ arr repeat >>>
		arr (parEvalN conf)) >>>
	app
\end{lstlisting}

\subsection{parMapStream}
\begin{center}
	\includegraphics[scale=0.7]{images/parMapStream}
\end{center}
As \code{parMap} is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}. So it makes sense to also have a \code{parMapStream} which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
\begin{lstlisting}[frame=htrbl]
parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> arr a b -> arr [a] [b]
parMapStream conf chunkSize f =
	(arr $ \as -> (f, as)) >>>
	(first $ arr repeat >>>
		arr (parEvalNLazy conf chunkSize)) >>>
	app
\end{lstlisting}

\subsection{farm}
\begin{center}
	\includegraphics[scale=0.7]{images/farm}
\end{center}
\code{parMap} spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper). This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
\begin{lstlisting}[frame=htrbl]
farm :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
	ArrowChoice arr, ArrowApply arr) =>
	conf -> NumCores -> arr a b -> arr [a] [b]
farm conf numCores f =
	(arr $ \as -> (f, as)) >>>
	(first $ arr mapArr >>> arr repeat >>>
		arr (parEvalN conf)) >>>
	(second $ arr (unshuffle numCores)) >>>
	app >>>
	arr shuffle
\end{lstlisting}
The definition of \code{unshuffle} is
\begin{lstlisting}[frame=htrbl]
unshuffle :: Int
	-> [a]
	-> [[a]]
unshuffle n xs = [takeEach n (drop i xs) | i <- [0..n-1]]
\end{lstlisting}
, while \code{shuffle} is defined as:
\begin{lstlisting}[frame=htrbl]
shuffle :: [[a]]
	-> [a]
shuffle = concat . transpose
\end{lstlisting}
(These were taken from Eden's source code. \cite{eden_skel_shuffle})

\subsection{farmChunk}
\begin{center}
	\includegraphics[scale=0.7]{images/farmChunk}
\end{center}
As \code{farm} is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict. So we define \code{farmChunk} which uses \code{parEvalNLazy} instead of \code{parEvalN} like this:
\begin{lstlisting}[frame=htrbl]
farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
	ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
farmChunk conf chunkSize numCores f =
	(arr $ \as -> (f, as)) >>>
	(first $ arr mapArr >>> arr repeat >>>
		arr (parEvalNLazy conf chunkSize)) >>>
	(second $ arr (unshuffle numCores)) >>>
	app >>>
	arr shuffle
\end{lstlisting}

\subsection{parMapReduce}

>>>>

\section{Basic Skeletons}

With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.

@@

### 2017-03-19 paper/mapSkeletons.tex
wip on paper. added pictures from the presentation.

\section{Map-based Skeletons}
With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.

\fixme{Extra Kapitel für die Basic Skeletons, vll den Syntactic Sugar dann danach machen statt erst nach den maps}

\subsection{parEvalNLazy}
\begin{center}
	\includegraphics[scale=0.7]{images/parEvalNLazy}
\end{center}
\code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows. Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g. \code{map (arr . (+)) [1..]} or just because we need the arrows evaluated in chunks. \code{parEvalNLazy} fixes this. It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{(arr \$ chunksOf chunkSize)}. These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}. The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
\begin{lstlisting}[frame=htrbl]
parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
parEvalNLazy conf chunkSize fs =
	arr (chunksOf chunkSize) >>>
	listApp fchunks >>>
	arr concat
	where
		fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
\end{lstlisting}

\subsection{parEval2}
\begin{center}
	\includegraphics[scale=0.7]{images/parEval2}
\end{center}
We have only talked about the paralellization arrows of the same type up until now. But sometimes we want to paralellize heterogenous types as well. For this, we introduce a helper combinator \code{arrMaybe} first, that converts an arrow \code{arr a b} to an arrow \code{arr (Maybe a) (Maybe b)}.
\begin{lstlisting}[frame=htrbl]
arrMaybe :: (ArrowApply arr) => (arr a b) -> arr (Maybe a) (Maybe b)
arrMaybe fn = (arr $ go) >>> app
	where 
		go Nothing = (arr $ \Nothing -> Nothing, Nothing)
		go (Just a) = ((arr $ \(Just x) -> (fn, x)) >>> app >>> arr Just, (Just a))
\end{lstlisting}
With this, we can now easily write \code{parEval2} which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)}. We start by converting both arrows with \code{arrMaybe}, combining them with \code{***} into a new arrow \code{arr (Maybe a, Maybe c) (Maybe b, Maybe d)}. This is then replicated twice and fed into parEvalN to get a \code{arr [(Maybe a, Maybe c)] [(Maybe b, Maybe d)]}. We can then apply this arrow to the input \code{[(Just a, Nothing), (Nothing, Just c)]} and then extract the resulting values with \code{fromJust} and the \code{!!} operator on lists in the last step.
\begin{lstlisting}[frame=htrbl]
parEval2 :: (ArrowParallel arr a b conf,
	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf,
	ArrowApply arr) =>
	conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
parEval2 conf f g =
	(arr $ \(a, c) -> (f_g, [(Just a, Nothing), (Nothing, Just c)])) >>>
	app >>>
	(arr $ \comb -> (fromJust (fst (comb !! 0)), fromJust (snd (comb !! 1))))
where
	f_g = parEvalN conf $ replicate 2 $ arrMaybe f *** arrMaybe g
\end{lstlisting}

\subsection{parMap}
\begin{center}
	\includegraphics[scale=0.7]{images/parMap}
\end{center}
\code{parMap} is probably the most common skeleton for parallel programs. We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}. 
Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf, ArrowApply arr) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f =
	(arr $ \as -> (f, as)) >>>
	(first $ arr repeat >>>
		arr (parEvalN conf)) >>>
	app
\end{lstlisting}

>>>>

\section{Map-based Skeletons}
\subsection{parMap}
\begin{center}
	\includegraphics[scale=0.7]{images/parMap}
\end{center}
\code{parMap} is probably the most common skeleton for parallel programs. We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}. 
Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf, ArrowApply arr) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f =
	(arr $ \as -> (f, as)) >>>
	(first $ arr repeat >>>
		arr (parEvalN conf)) >>>
	app
\end{lstlisting}

### 2017-03-19 paper/parallelHaskells.tex
wip on paper. added pictures from the presentation.

\section{Short introduction to parallel Haskells}
There are already several ways to write parallel programs in Haskell. As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
\begin{center}
	\includegraphics[scale=0.7]{images/parEvalN}
\end{center}
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
We will now implement \code{parEvalN} with the different parallel Haskells.

\subsection{Multicore Haskell}
Multicore Haskell \cite{multicore_hackage} is the GHC native way to do parallel processing. It ships with parallel evaluation strategies for several types which can be applied with \code{using :: a -> Strategy a -> a}. For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}. This lazy list \code{[b]} is then forcibly evaluated in parallel with the strategy \code{Strategy [b]} by the \code{using} operator. This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a}.
\begin{center}
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
\end{center}
\begin{lstlisting}[frame=htrbl]
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = zipWith ($) fs as `using` parList rdeepseq
\end{lstlisting}

\subsection{ParMonad}
The \code{Par} monad introduced by Marlow et al. \cite{monad_par_paper_2011}, which can be found in the monad-par package on hackage \cite{monad_par_hackage}, is a monad designed for composition of parallel programs.
\\\\
Our parallel evaluation function \code{parEvalN} can be defined by zipping the list of \code{[a -> b]} with the list of inputs \code{[a]} with the application operator \code{\$} just like with Multicore Haskell. Then, we map over this not yet evaluated lazy list of results \code{[b]} with \code{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \code{[Par (IVar b)]}, which we convert to \code{Par [IVar b]} with \code{sequenceA}. We wait for the computations to finish by mapping over the \code{IVar b}'s inside the \code{Par} monad with \code{get}. This results in \code{Par [b]}. We finally execute this process with \code{runPar} to finally get \code{[b]} again.
\begin{center}
	\includegraphics[scale=0.5]{images/parEvalNParMonad}
\end{center}
\begin{lstlisting}[frame=htrbl]
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = runPar $ 
	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
\end{lstlisting}

\subsection{Eden}
Eden  is a parallel Haskell for distributed memory and comes with a MPI and a PVM backend \cite{eden_hackage, Loogen2012, eden_homepage}. This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell.
\\\\
While it also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
\\
\code{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
\\
This allows us to define \code{parEvalN} quite easily:
\begin{center}
	\includegraphics[scale=0.5]{images/parEvalNEden}
\end{center}
\begin{lstlisting}[frame=htrbl]
parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
parEvalN fs as = spawnF fs as
\end{lstlisting}

>>>>

\section{Short introduction to parallel Haskells}
There are already several ways to write parallel programs in Haskell. As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:

\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
\begin{center}
	\includegraphics[scale=0.7]{images/parEvalN}
\end{center}
We will now implement \code{parEvalN} with the different parallel Haskells.

\subsection{Multicore Haskell}
Multicore Haskell \cite{multicore_hackage} is the GHC native way to do parallel processing. It ships with parallel evaluation strategies for several types which can be applied with \code{using :: a -> Strategy a -> a}. For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}. This lazy list \code{[b]} is then forcibly evaluated in parallel with the strategy \code{Strategy [b]} by the \code{using} operator. This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a}.

\begin{lstlisting}[frame=htrbl]
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = zipWith ($) fs as `using` parList rdeepseq
\end{lstlisting}
\begin{center}
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
\end{center}

\subsection{ParMonad}
The \code{Par} monad introduced by Marlow et al. \cite{monad_par_paper_2011}, which can be found in the monad-par package on hackage \cite{monad_par_hackage}, is a monad designed for composition of parallel programs.
\\\\
Our parallel evaluation function \code{parEvalN} can be defined by zipping the list of \code{[a -> b]} with the list of inputs \code{[a]} with the application operator \code{\$} just like with Multicore Haskell. Then, we map over this not yet evaluated lazy list of results \code{[b]} with \code{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \code{[Par (IVar b)]}, which we convert to \code{Par [IVar b]} with \code{sequenceA}. We wait for the computations to finish by mapping over the \code{IVar b}'s inside the \code{Par} monad with \code{get}. This results in \code{Par [b]}. We finally execute this process with \code{runPar} to finally get \code{[b]} again.

\begin{lstlisting}[frame=htrbl]
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = runPar $ 
	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
\end{lstlisting}
\begin{center}
	\includegraphics[scale=0.5]{images/parEvalNParMonad}
\end{center}

\subsection{Eden}
Eden  is a parallel Haskell for distributed memory and comes with a MPI and a PVM backend \cite{eden_hackage, Loogen2012, eden_homepage}. This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell.
\\\\
While it also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
\\
\code{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
\\
This allows us to define \code{parEvalN} quite easily:

\begin{lstlisting}[frame=htrbl]
parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
parEvalN fs as = spawnF fs as
\end{lstlisting}
\begin{center}
	\includegraphics[scale=0.5]{images/parEvalNEden}
\end{center}

### 2017-03-19 paper/parrows.tex
wip on the paper
wip on paper. added pictures from the presentation.

\section{Parallel Arrows}
We have seen what Arrows are and how they can be used as a general interface to computation. In the following section we will discuss how Arrows can also be looked at as a general interface not only to computation, but to \textbf{parallel computation} as well. We start by introducing the interface and explaining the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\subsection{The ArrowParallel typeclass}
As we have seen earlier, in its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function listApp, but does parallel instead of serial evaluation).
\begin{lstlisting}[frame=htrbl]
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{lstlisting}
With this definition of \code{parEvalN}, parallel execution can be looked at as yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{lstlisting}
Sometimes parallel Haskells require additional configuration parameters as information about the execution environment. This is why we also introduce an additional \code{conf} parameter to the function. We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}. So we add it to the type signature of the typeclass as well.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{lstlisting}
We don't require the conf parameter in every implementation. If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN} as we will see in the implementation of the Multicore and the ParMonad backend.

@@

\subsection{Benefits of parallel Arrows}
We have seen that we can wrap parallel Haskells inside of the \code{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
\begin{itemize}
	\item \textbf{Arrow API benefits}:
	With the \code{ArrowParallel} typeclass we do not lose any benefits of using arrows as \code{parEvalN} is just yet another arrow combinator. The resulting arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types. We can just "plug" in parallel parts into our sequential programs without having to change anything.
	\item \textbf{Abstraction}:
	With the \code{ArrowParallel} typeclass, we abstracted all parallel implementation logic away from the business logic. This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell. So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just swapping out the actual \code{ArrowParallel} instance.
\end{itemize}

>>>>

\section{Parallel Arrows}
We have seen what Arrows are and how they can be used as a general interface to computation. In the following section we will discuss how Arrows can also be looked at as a general interface not only to computation, but to \textbf{parallel computation} as well. We start by introducing the interface and explaining the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\begin{center}
	\includegraphics[scale=0.7]{images/parEvalN}
\end{center}
\subsection{The ArrowParallel typeclass}
As we have seen earlier, in its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function listApp, but does parallel instead of serial evaluation).
\begin{lstlisting}[frame=htrbl]
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{lstlisting}
With this definition of \code{parEvalN}, parallel execution can be looked at as yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{lstlisting}
Sometimes parallel Haskells require additional configuration parameters as information about the execution environment. This is why we also introduce an additional \code{conf} parameter to the function. We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}. So we add it to the type signature of the typeclass as well.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{lstlisting}
We don't require the conf parameter in every implementation. If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN} as we will see in the implementation of the Multicore and the ParMonad backend.

@@

\subsection{Benefits of parallel Arrows}
\fixme{Okay so?}
\\
We have seen that we can wrap parallel Haskells inside of the \code{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
\begin{itemize}
	\item \textbf{Arrow API benefits}:
	With the \code{ArrowParallel} typeclass we do not lose any benefits of using arrows as \code{parEvalN} is just yet another arrow combinator. The resulting arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types. We can just "plug" in parallel parts into our sequential programs without having to change anything.
	\item \textbf{Abstraction}:
	With the \code{ArrowParallel} typeclass, we abstracted all parallel implementation logic away from the business logic. This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell. So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just swapping out the actual \code{ArrowParallel} instance.
\end{itemize}
